Orchestrating stencils with the Multi-Stencil Framework

### 2017-03-02 ARTICLES/IJPP/articleR1/introduction.tex
intro

As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. For example, the current world's top supercomputer Tianhe-2~\footnote{\url{www.top500.org}} is composed of multi-cores processors and accelerators, and is able to reach a theoretical peak performance around thirty peta floating point operations par seconds. However, to be able to use such a machine, multiple programming models, such as MPI (Message Passing Interface), OpenMP, CUDA etc., and multiple optimization techniques, such as cache or vectorization optimizations, have to be combined. Moreover, current architectures evolution lets think that heterogeneity and complexity in HPC will continue to grow in future.

@@

However, many domain specific languages and frameworks have been proposed for many domains yet, and each one claims to solve a specific case that others don't. Furthermore instead of re-using a common basis for every language or framework of a same domain, each solution is implemented from scratch as a monolithic code. We claim in this paper that being able to find a common meta-model for all solutions of a given domain facilitates code re-use, and eases flexibility and maintainability of solutions.

For example, to numerically solve a set of partial differential equations (PDEs), iterative methods are frequently used to approximate the exact solution through a discretized phenomena. A very well known and usual way to discretize PDEs is to transform them to explicit numerical schemes, also often called \emph{stencils}. Many DSLs have been proposed for stencil computations~\cite{spaaTangCKLL11,citeulike12258902,Ragan-Kelley:2013:HLC:2491956.2462176,DeVito:2011:LDS:2063384.2063396,Camier:2015:IPP:2820083.2820107}, as it will be detailed in Section~\ref{sect:rel}. Many of them use same kind of parallelization, data structures or optimization, however each one has been built from scratch to deal with another additional specific case.

We propose the Multi-Stencil Framework (MSF) that is built over a meta-model of multi-stencil simulations. MSF produces a parallel orchestration of a multi-stencil program without being aware of the underlying implementation choices (\eg distributed data structures, task scheduler etc.). Thanks to this meta-model MSF is able to easily switch from one parallelization technique to another and from one optimization to another. Moreover, as the meta-model is independent from implementation details, MSF can easily choose one back-end or another, thus easing code reuse of existing solutions.

Contributions of the paper are: a meta-model of a multi-stencil program presented in Section~\ref{}; from this meta-model are built both a light and descriptive domain specific language MSL as well as a generic template of the application described in Section~\ref{}; the compiler MSC of the framework builds a parallel specialized version of the generic template from the specific application described with MSL; finally a performance evaluation is detailed in Section~\ref{}  and validates the performance model introduced in Section~\ref{}.

>>>>

As the computation power of modern high performance architectures increases, their heterogeneity and complexity also become more important. For example, the current world's top supercomputer Tianhe-2~\footnote{\url{www.top500.org}} is composed of multi-cores processors and accelerators, and is able to reach a theoretical peak performance around thirty peta floating point operations par seconds. However, to be able to use such a machine, multiple programming models, such as MPI (Message Passing Interface), OpenMP, CUDA etc., and multiple optimization techniques, such as cache optimization, have to be combined. Moreover, current architectures evolution lets think that heterogeneity and complexity in HPC will continue to grow in future.

@@

Many domain specific languages and frameworks have been proposed for many domains yet, and each one claims to solve a specific case that others don't. However, instead of re-using a common basis for every language or framework of a same domain, each solution is implemented from scratch as a monolithic code. We claim in this paper that being able to find a common meta-model to all solutions for a given domain facilitates code re-use, and eases flexibility and maintainability of solutions.

For example, to numerically solve a set of partial differential equations (PDEs), iterative methods are frequently used to approximate the exact solution through a discretized phenomena. A very well known and usual way to discretize PDEs is to transform them to explicit numerical schemes, also often called \emph{stencils}. Many DSLs have been proposed for stencil computations~\cite{spaaTangCKLL11,citeulike12258902,Ragan-Kelley:2013:HLC:2491956.2462176,DeVito:2011:LDS:2063384.2063396,Camier:2015:IPP:2820083.2820107}, as it will be detailed in Section~\ref{sect:rel}. Many of them use same kind of parallelization, data structures or optimization techniques, however each one has been built from scratch.

We propose the Multi-Stencil Framework (MSF) that is built over a meta-model of multi-stencil simulations. MSF produces a parallel orchestration of a multi-stencil program without being aware of the underlying implementation choices (\eg distributed data structures, task scheduler etc.). Thanks to this meta-model MSF is able to easily switch from one parallelization technique to another and from one optimization to another. Moreover, as the meta-model is independent from implementation details, MSF can easily choose one back-end or another, thus easing code reuse of existing solutions. To ease composition of existing solutions, MSF uses Software Engineering capabilities of component-based programming models~\cite{Szyperski:2002:CSB:515228} where applications are defined as an assembly of building blocks, or components.

After a short overview of the Multi-Stencil Framework given in Section~\ref{sect:msf}, contributions of the paper are presented as follows: a meta-model of a multi-stencil program presented in Section~\ref{sect:formalism}; from this meta-model are built both a light and descriptive domain specific language, namely MSL, as well as a generic component assembly of the application described in Section~\ref{sect:msl}; the compiler MSC of the framework builds a parallel specialized version of the generic assembly from the application description with MSL, MSC and its performance model are described in Section~\ref{sect:parallelism}; finally a performance evaluation is detailed in Section~\ref{sect:eval}  and validates the performance model.
